---
title: "Getting started with Rfishmorph"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with Rfishmorph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
```

```{r setup}
library(Rfishmorph)
```

## The FISHMORPH scheme

`Rfishmorph` implements the FISHMORPH morphospace of Brosse et al. (2021): 21
landmarks digitized on a lateral photograph, from which 11 body segments and 9
dimensionless ratios are derived.

```{r}
sc <- fishmorph_schema()
sc$segments
sc$ratios
data.frame(ratio = sc$ratios,
           definition = vapply(sc$ratio_def, paste, "", collapse = " / "),
           `function` = sc$ratio_function[sc$ratios],
           check.names = FALSE)
```

## From landmarks to traits

```{r}
lm <- read_landmarks_csv(
  system.file("extdata", "example_landmarks.csv", package = "Rfishmorph"))
lm

seg <- fishmorph_segments(lm, scale_cm = 1)
seg

fishmorph_ratios(seg)
```

```{r, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot_landmarks(lm, specimen = 1)
```

## Reconstruction and the round-trip guarantee

`reconstruct_fishmorph_landmarks()` inverts `fishmorph_segments()`. The free
parameters move landmarks around but never change segment lengths, so the
round-trip is exact.

```{r}
seg1 <- list(Bl = 10, Bd = 3.2, Hd = 2.4, Eh = 1.1, Mo = 1.6, PFi = 1.9,
             PFl = 2.1, Ed = 0.6, Jl = 1.3, CPd = 1.0, CFd = 3.0)
check_reconstruction_roundtrip(seg1)
```

## The functional trait space

The ordination is fitted once on the reference database and then frozen, so new
specimens are *projected* into the same coordinate system rather than
re-estimating the axes.

```{r}
ref <- load_fishmorph_reference()
ts  <- fishmorph_trait_space(ref, groups = ref$Order)
ts
```

```{r}
plot_trait_space(ts, style = "hull")                 # per-group convex hulls
plot_trait_space(ts, style = "hull", reference_density = TRUE)  # + global density
```

## Quality control: segments vs landmarks

Two scale-invariant quantities let you check a workbook's internal consistency:
the 9 ratios and the segments expressed as a fraction of `Bl`.

```{r}
segtab <- data.frame(
  Genus.species = paste0("sp", 1:6), Bl = 10,
  Bd = c(3, 3.2, 3.4, 3.6, 3.8, 4), Hd = c(2, 2.1, 2.2, 2.3, 2.4, 2.5),
  Eh = 1, Mo = 1.5, PFi = 2, PFl = 2.5, Ed = 0.5, Jl = 1.2, CPd = 1, CFd = 3)
cmp <- compare_segments_landmarks(segtab, segtab, id_col = "Genus.species")
summary(cmp)
```

## Adding a new species

```{r}
rec <- new_fishmorph_species(
  "Genus novus",
  segments = list(Bl = 10, Bd = 3.2, Hd = 2.4, Eh = 1.1, Mo = 1.6, PFi = 1.9,
                  PFl = 2.1, Ed = 0.6, Jl = 1.3, CPd = 1.0, CFd = 3.0),
  family = "Cyprinidae", order = "Cypriniformes")
ref2 <- add_fishmorph_species(ref, rec)
nrow(ref2)
```

Name resolution and the global freshwater list require `rfishbase` and network
access:

```{r, eval = FALSE}
validate_species_names(c("Salmo trutta", "Esox_lucius"))
fw <- freshwater_fish_list()
```

## Reference

Brosse et al. (2021) *Global Ecology and Biogeography* 30:2330–2336.
<https://doi.org/10.1111/geb.13395>
